---
title: "Getting started with networkGen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with networkGen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`networkGen` builds a PCSF (prize-collecting Steiner forest) network from
kinase-activity data, optionally paired with sensitivity data. This vignette
covers: a tiny, hand-verifiable toy example; PCSF's key parameters; and
`run_network_grid()`, the one entry point for real, save-worthy output --
whether that's one network or many, kinase-only or paired, one parameter
combination or a whole grid of them.

```{r setup}
library(networkGen)
```

## 1. A tiny, verifiable toy example

Terminal nodes `A`, `B`, `C` need connecting. Two candidate paths exist
between `B` and `C`: a cheap one through `D` (total cost 0.2), and a more
expensive one through `E` and `F` (total cost 0.3). PCSF should pick the
cheap path -- `D` should appear in the result, `E` and `F` should not.

```{r toy-data}
toy_ppi <- data.frame(
  head = c("A", "B", "D", "B", "E", "F"),
  tail = c("B", "D", "C", "E", "F", "C"),
  cost = c(0.1, 0.1, 0.1, 0.1, 0.1, 0.1)
)
toy_uka <- data.frame(
  name = c("A", "B", "C"), prize = c(0.9, 0.9, 0.9),
  type = "Kinase", LogFC = c(1, 1, 1)
)
```

```{r toy-build}
toy_result <- generate_kinase_network(
  uka = toy_uka, ppi_network = toy_ppi,
  spec_cutoff = 0, condition = "toy_example", seed = 12345, n = 2
)
cat(paste(toy_result$edges$from, "-", toy_result$edges$to), sep = "\n")
```

`A - B`, `B - D`, `D - C` -- the cheap path through `D`. `E` and `F` never
appear.

## 2. How PCSF builds a network, and its two main knobs

PCSF (prize-collecting Steiner forest) starts from a reference network whose
**nodes carry prizes** -- here, each input kinase's percentile rank -- and
whose **edges carry costs**. It picks a subnetwork that trades those off,
roughly:

```
minimize (  sum of edge costs  +  b * sum of prizes of input nodes left out  )
```

Pulling an input node in costs edge weight; leaving it out forfeits `b`
times its prize. Two parameters set where that balance lands:

- **`b`** -- multiplies every prize. Higher `b` makes leaving nodes out
  expensive, so PCSF pulls in more input nodes **and** more connecting
  (Steiner) nodes: a bigger network. Default `2`.
- **`w`** -- the cost of attaching an input node straight to an artificial
  root instead of routing to it through the network. Low `w` lets isolated
  nodes in cheaply (they show up as tiny fragments, or drop out of the node
  list entirely); high `w` means a node is only kept if the network already
  runs near it -- fewer, better-connected pieces. A lone node survives only
  when `b * prize > w`. Default `2` (PCSF's own default).

If PCSF reports it can't find a subnetwork, raise `b` (or lower `w`). The
defaults suit input node sets that are biologically related (real kinase
hits from one comparison usually are); an unrelated set may need a larger
`b` and `w` to knit together at all -- so this vignette leaves `b`/`w` at
their defaults everywhere and the examples still build real networks.

Less commonly tuned: `n` (randomized runs combined into one consensus
network -- default 8, lowered to `n = 2` throughout this vignette for
speed), `r` (per-run random noise fraction), `mu` (penalty on routing
through high-degree hub genes).

Full parameter documentation: `` ?generate_kinase_network ``.

`networkGen` also ships a real reference PPI network as package data --
`ppi_networkv12` (the default `ppi_network`, used automatically if you don't
pass one), plus two narrower variants: `ppi_networkv12_filt`, and
`ppi_networkv12_502_kins` (the smallest -- restricted to 502 kinases, ~107K
edges vs. 1.1M for the full network -- good enough for a kinase-activity
network and much faster, used throughout the rest of this vignette):

```{r ref-network}
nrow(ppi_networkv12)
nrow(ppi_networkv12_502_kins)
head(ppi_networkv12_502_kins)
```

## 3. Running multiple networks, optionally in parallel

`run_network_grid()` is the one entry point for this package -- one network
or many, kinase-only or paired, one parameter combination or a whole grid of
them. You specify the parameters to tune (e.g. cutoffs, `b`, `w`,
`ppi_network`, which datasets) and `run_network_grid()` builds a grid of
every combination.

For each condition and parameter combination, `run_network_grid()`:

1. filters the kinase-activity data to its top hits (`spec_cutoff` on
   specificity, `perc_cutoff` on percentile rank)
2. creates one output folder, with `params.csv` recording exactly what
   parameters produced it
3. runs the network and writes its results into that same folder

Networks run in parallel if a `future` plan is set (see the end of this
section); sequentially otherwise.

### Multiple datasets, multiple `spec_cutoff` and `b` values

UKA exports are named `UKA_<PTK|STK>_<number>_<dataset name>.csv`. 3 real
timepoints below, each a real PTK+STK pair:

```{r multidataset-load}
multidataset_paths <- list.files(
  "example_data/multidataset", pattern = "\\.csv$", recursive = TRUE, full.names = TRUE
)
basename(multidataset_paths)

raw_multidataset <- load_uka_dataset_files(multidataset_paths)
table(raw_multidataset$dataset, raw_multidataset$kinase_type)
```

This is raw (not pre-cleaned) data, so `run_network_grid()` needs a
`clean_fn` -- `Sgroup_contrast` is the condition column
`clean_uka_to_kinograte()` produces. `spec_cutoff` **and** `w` as vectors
grid every dataset's condition against every cutoff x `w` combination --
3 datasets x 2 `spec_cutoff` values x 2 `w` values below, no loop written
(`perc_cutoff`, `b`, `rank_uka_abs` and `ppi_network` could grid the same
way; kept at their defaults here so the example stays readable):

```{r multidataset-build}
multidataset_dir <- tempfile()
dir.create(multidataset_dir)

multidataset_results <- run_network_grid(
  raw_multidataset, clean_fn = clean_uka_to_kinograte, condition_col = "Sgroup_contrast",
  spec_cutoff = c(1, 1.5), perc_cutoff = 0.5, w = c(2, 4), ppi_network = ppi_networkv12_502_kins,
  respath = multidataset_dir, write = TRUE, n = 2
)

for (item in multidataset_results) {
  cat(item$meta$dataset, "|", item$meta$condition, "| spec_cutoff =", item$meta$spec_cutoff,
    "| w =", item$meta$w, ": ",
    if (is.null(item$result)) "no subnetwork" else paste(nrow(item$result$nodes), "nodes"), "\n",
    sep = ""
  )
}

list.files(multidataset_dir, recursive = TRUE)
```

Four folders (one per `spec_cutoff` x `w` combination), each holding all 3
datasets' `nodes`/`edges`/`wc_df` CSVs plus one shared `params.csv` -- not
one folder per (dataset, `spec_cutoff`, `w`) combination, and not one folder
covering multiple `spec_cutoff`/`w` values at once.

### Paired (with sensitivity data), multiple `perc_cutoff` values on each side

The same drug's 3 doses, paired with that cell line's (`NUDUL1`) real
sensitivity profile -- one sensitivity profile, several kinase-activity
conditions from the same cell line. Passing raw `sens` data (not
pre-filtered) plus `sens_perc_cutoff` switches `run_network_grid()` to build
paired (kinase + sensitivity) networks instead of kinase-only, filtering
that one sensitivity profile per `sens_perc_cutoff` value and reusing each
filtered version across every condition. `perc_cutoff` (kinase side) and
`sens_perc_cutoff` (sensitivity side) grid independently of each other --
there's no `spec_cutoff`-equivalent for sensitivity data, since specificity
is a kinase-activity-only concept:

```{r batch-paired-clean}
uka_doses <- read.csv("example_data/cdk_inhibitor_PHA-793887_all_doses.csv")
unique(uka_doses$Sgroup_contrast)

sens <- read.csv("example_data/sensitivity_example.csv")
sens_nudul <- sens[sens$cell_line == "NUDUL1", ]
```

```{r batch-paired-build}
paired_dir <- tempfile()
dir.create(paired_dir)

paired_results <- run_network_grid(
  uka_doses, condition_col = "Sgroup_contrast",
  spec_cutoff = 0.5, perc_cutoff = 0.5,
  sens = sens_nudul, sens_perc_cutoff = c(0.6, 0.7),
  ppi_network = ppi_networkv12_502_kins, respath = paired_dir, write = TRUE, n = 2
)

for (item in paired_results) {
  cat(item$meta$condition, "| sens_perc_cutoff =", item$meta$sens_perc_cutoff, ": ",
    nrow(item$result$nodes), " nodes, ", nrow(item$result$edges), " edges\n",
    sep = ""
  )
}

list.files(paired_dir, recursive = TRUE)
```

The folder additionally nests under a `uka`-named subfolder (since output
there depends on both `uka` and `sens`), with one subfolder per distinct
`sens_perc_cutoff` value inside it (2 here, since `perc_cutoff`/`spec_cutoff`
are each a single value in this example).

### Safety guard for large grids

`run_network_grid()` refuses to build more than `max_tasks` (default 500)
networks without an explicit override -- a check against an accidentally
huge overnight run, not a time estimate:

```{r max-tasks, error = TRUE}
run_network_grid(
  uka_doses, condition_col = "Sgroup_contrast",
  spec_cutoff = seq(0.1, 0.9, by = 0.1), perc_cutoff = 0.7,
  respath = paired_dir, write = FALSE, max_tasks = 5
)
```

### Parallel vs. sequential execution

By default, no `future` plan is set, so `run_network_grid()` runs
**sequentially** -- one build at a time, still correct, just not parallel.
To actually parallelize across the whole grid, set a plan **once**, before
any `run_network_grid()` call in the session (not per call -- worker
startup, including loading `PCSF`'s compiled code in each worker, has real
cost, so it's not worth repeating):

```{r batch-plan, eval = FALSE}
future::plan(future::multisession, workers = 4) # local parallel workers
```

Every `run_network_grid()` call for the rest of the session now runs in
parallel automatically -- nothing else about the call changes. To go back
to sequential:

```{r batch-plan-reset, eval = FALSE}
future::plan(future::sequential)
```

Persisted output is the `nodes`/`edges`/`wc_df` CSVs written above, plus
`params.csv` -- that's what `networkPlot` (enrichment + visualization) reads
back in later, not a saved copy of the `igraph` object itself.
